Exploring the evolution of stellar rotation using Galactic kinematics

### 2020-04-16 paper/ms63.tex
moving the table to the appendix
adding some changes from the referee report

\begin{abstract}
The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
In this work, we use the velocity dispersions of groups of low-mass \kepler\
dwarfs as an age proxy, to reveal the temperature-dependent rotational
evolution of these stars, and demonstrate that kinematic ages could be a
useful tool for calibrating gyrochronology in future.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K and late-G dwarfs when
they are young, at old ages we find that late-G and early-K dwarfs rotate at
the {\it same rate} or faster than late-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
GKM dwarfs are likely to be synchronized binaries.
\end{abstract}

@@

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog of parallaxes, proper motions and apparent magnitudes.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.

For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
To calibrate this relation, Curtis \etal\ (2020, in prep) combined effective
temperature measurements for nearby, unreddened field stars in benchmark
samples, including FGK stars characterized with high-resolution optical
spectroscopy \citep{brewer2016}, M dwarfs characterized with low-resolution
optical and near-infrared spectroscopy \citep{mann2015}, and K and M dwarfs
characterized with interferometry and bolometric flux analyses
\citep{boyajian2012}.
This empirical color--temperature relation is valid over the color range $0.55
< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $6470 < T_{\rm eff} <
3070$~K.
The dispersion about the relation implies a high precision of 50~K.
These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
at all temperatures\footnote{The color--temperature relation is described in
detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
\etal\ (2020, in prep).}.
The equation we used to calculate photometric temperatures from Gaia \gcolor\
color is a seventh-order polynomial with coefficients given in table
\ref{tab:coeffs}.
\begin{table}[h!]
  \begin{center}
      \caption{
          Coefficient values for the 7th-order polynomial used to estimate
      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
      prep).}
    \label{tab:coeffs}
    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
        (\gcolor\ ) exponent & Coefficient  \\
      \hline
      $0$ & -416.585 \\
      $1$ & 39780.0  \\
      $2$ & -84190.5 \\
      $3$ & 85203.9  \\
      $4$ & -48225.9 \\
      $5$ & 15598.5  \\
      $6$ & -2694.76 \\
      $7$ & 192.865  \\
    \end{tabular}
  \end{center}
\end{table}

@@

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
gyrochrones from the \citet{spada2019} model shown in white.
These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
Each point plotted in the top panel also appears in the bottom panel with the
same color.
Because velocity dispersion was calculated in bins of \teff, not mass, bin
outlines are clearly visible in the top panel but appear smeared-out in the
bottom panel.
In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
{\it do} trace lines of constant velocity dispersion, and reproduce the trends
in the data at all ages.
These models qualitatively agree with the data and reproduce the apparent
flattening and inversion in the rotation period-\teff/mass relations.

@@

The velocity dispersions of stars in the \mct\ sample, shown in figure
\ref{fig:vplot}, provide the following picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010}.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

@@

\section{Appendix: Validating \vb\ dispersion as an age proxy}
\label{sec:mass-dependent-heating}

@@

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyr and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because it only
includes stars with {\it detectable} rotation periods, and since lower-mass
stars stay active for longer it is likely that it contains a strong mass-age
correlation.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $<$ 4.
We also eliminated photometric binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
figure \ref{fig:age_gradient}).
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
outliers before calculating the standard deviation of stars in each bin.
These extreme velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

@@

>>>>

\begin{abstract}
The rotational evolution of cool dwarfs is poorly constrained after $\sim$1-2
Gyr due to a lack of precise ages and rotation periods for old main-sequence
stars.
In this work, we use the velocity dispersions of groups of low-mass \kepler\
dwarfs as an age proxy, to reveal the temperature-dependent rotational
evolution of these stars, and demonstrate that kinematic ages could be a
useful tool for calibrating gyrochronology in future.
We find that a linear gyrochronology model, calibrated to fit the
period--\teff\ relationship of the Praesepe cluster, does not apply to stars
older than around 1 Gyr.
Although late-K dwarfs spin more slowly than early-K dwarfs when they are
    young, at old ages we find that late-K dwarfs rotate at the {\it same
    rate} or faster than early-K dwarfs of the same age.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that the surface rotation rates of K dwarfs go through an epoch
of inhibited evolution.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we confirm that many rapidly rotating
GKM dwarfs are likely to be synchronized binaries.
\end{abstract}

@@

We used the publicly available \kepler-\gaia\ DR2 crossmatched
catalog\footnote{Available at gaia-kepler.fun} to combine the \mct\ catalog of
stellar rotation periods, measured from \kepler\ light curves, with the \gaia\
DR2 catalog of parallaxes, proper motions and apparent magnitudes.
Reddening and extinction from dust was calculated for each star using the
Bayestar dust map implemented in the {\tt dustmaps} {\it Python} package
\citep{green2018}, and {\tt astropy} \citep{astropy2013, astropy2018}.
For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
\racomment{The calibration of this relation is described Curtis \etal\ (2020,
in prep) and briefly summarized in the Appendix of this paper.}

@@

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
gyrochrones from the \citet{spada2019} model shown in white.
These gyrochrones are \racomment{plotted for the same ages as above.}
Each point plotted in the top panel also appears in the bottom panel with the
same color.
Because velocity dispersion was calculated in bins of \teff, not mass, bin
outlines are clearly visible in the top panel but appear smeared-out in the
bottom panel.
In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
{\it do} trace lines of constant velocity dispersion, and reproduce the trends
in the data at all ages.
These models qualitatively agree with the data and reproduce the apparent
flattening and inversion in the rotation period-\teff/mass relations.

@@

\racomment{
In the top panel of figure \ref{fig:vplot}, the \citet{angus2019}
gyrochronology model is over-plotted for comparison.
We used this model as an example, however all empirical gyrochronology models
are similar in essence to the \citet{angus2019} relation and none is able to
reproduce the observed velocity dispersions, or capture the evolving shape of
the period-\teff\ relations.
The \citet{angus2019} relation is a new fully-empirical gyrochronology
relation, calibrated using recently measured rotation periods for members of
the Praesepe cluster, which extend down to early M dwarfs \citep{rebull2017,
douglas2017}.
These are the oldest cluster M dwarfs with measured photometric rotation
periods and the \citet{angus2019} model therefore encapsulates the behavior of
these cool stars at Praesepe age.

Alternatively, we could have chosen to compare our data with one of the many
other available empirical gyrochronology models.
For example, the \citet{barnes2003, barnes2007, meibom2011, angus2015}
relations all have a functional form that was first introduced by
\citet{barnes2003}: $\mathrm{P_{rot}} = t^n~a(B-V - c)^b$, where n, a, b and c
are free parameters, and B and V are photometric magnitudes.
These relations have been widely used in the literature, and are similar to
the \citet{angus2019} relation in that they consist of a color-dependent term
multiplied by an age-dependent term, \ie\ they are separable in color and age.
None of these relations follow lines of constant velocity dispersion in figure
\ref{fig:vplot} because they have a universal power-law index, $n$ (around
0.5), which applies to stars of all masses and ages.
They do not have the flexibility to capture the evolving period-color
relationship seen in the data.
The \citet{barnes2010} model has a different functional form: it uses Rossby
number ($Ro = P_\mathrm{rot}/\tau$, where $\tau$ is the convective turnover
time) to encode the mass-dependent evolution of rotation periods:
$\frac{dP_\mathrm{rot}}{dt} \propto 1/Ro$.
This simple relation neatly reproduces the rotation periods of young cluster
stars, however, like the other empirical gyrochronology models, it does not
reproduce the observed trends seen in the velocity dispersion data.
It has a similar period-\teff\ relation to the Praesepe-based
\citet{angus2019} model, because the period-\teff\ relation of Praesepe
roughly follows a line of constant Rossby number.
Consequently, like the \citet{angus2019} model, the \citet{barnes2010}
gyrochrones do not reproduce the data explored here and do not have the
flexibility to capture mass and time-variable rotational evolution.
Although it seems that the period-\teff\ relations follow lines of constant
}

The velocity dispersions of stars in the \mct\ sample, shown in figure
\ref{fig:vplot}, provide the following picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `C' sequence to the `I'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, kraft1967,
parker1970, kawaler1988, charbonneau2010}.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the \citet{spada2019} model, the radiative cores and convective
envelopes of stars are decoupled at these young ages, \ie\ transportation of
angular momentum from the surface to the core of the star is reduced, so the
surface slows down due to wind-braking but the core keeps spinning rapidly.
According to the data presented in figure \ref{fig:vplot}, at intermediate
ages, the rotation periods of K dwarfs appear {\it constant} with mass, and at
late ages rotation period {\it increases} with {\it increasing} mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

@@

\section{Appendix A: Validating \vb\ dispersion as an age proxy}
\label{sec:mass-dependent-heating}

@@

To investigate whether mass-dependent heating could be acting on the \kepler\
sample, we selected late K and M dwarfs observed by both \kepler\ and \gaia,
whose MS lifetimes exceed around 11 Gyr and are therefore representative of
the initial mass function.
We could not perform this analysis on the \mct\ sample, because it only
includes stars with {\it detectable} rotation periods, and since lower-mass
stars stay active for longer it is likely that it contains a strong mass-age
correlation.
We selected all \kepler\ targets with dereddened \gaia\ \gcolor\ colors
greater than 1.2 (corresponding to an effective temperature $\lesssim$
4800 K) and absolute \gaia\ $G$-band magnitudes $>$ 4.
We also eliminated photometric binaries by removing stars above a 6th order
polynomial, fit to the MS on the \gaia\ CMD (similar to the one shown in
figure \ref{fig:age_gradient}).
We then applied the quality cuts described above in section
\ref{sec:the_data}.
To search for evidence of mass-dependent heating we calculated the (\vb)
velocity dispersion of stars in effective temperature bins.
Sigma clipping was performed at 3$\sigma$ to remove high and low velocity
outliers before calculating the standard deviation of stars in each bin.
These extreme velocity outliers may be very old late K and M dwarfs, or they
result from using \vb\ instead of \vz, which introduces additional velocity
scatter.

@@

\racomment{
\section{Appendix B: Photometric temperatures}

For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
To calibrate this relation, Curtis \etal\ (2020, in prep) combined effective
temperature measurements for nearby, unreddened field stars in benchmark
samples, including FGK stars characterized with high-resolution optical
spectroscopy \citep{brewer2016}, M dwarfs characterized with low-resolution
optical and near-infrared spectroscopy \citep{mann2015}, and K and M dwarfs
characterized with interferometry and bolometric flux analyses
\citep{boyajian2012}.
This empirical color--temperature relation is valid over the color range $0.55
< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $3070 > T_{\rm eff} >
6470$~K.
The dispersion about the relation implies a high precision of 50~K.
These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
at all temperatures\footnote{The color--temperature relation is described in
detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
\etal\ (2020, in prep).}.
The equation we used to calculate photometric temperatures from Gaia \gcolor\
color is a seventh-order polynomial with coefficients given in table
\ref{tab:coeffs}.
\begin{table}[h!]
  \begin{center}
      \caption{
          Coefficient values for the 7th-order polynomial used to estimate
      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
      prep).}
    \label{tab:coeffs}
    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
        (\gcolor\ ) exponent & Coefficient  \\
      \hline
      $0$ & -416.585 \\
      $1$ & 39780.0  \\
      $2$ & -84190.5 \\
      $3$ & 85203.9  \\
      $4$ & -48225.9 \\
      $5$ & 15598.5  \\
      $6$ & -2694.76 \\
      $7$ & 192.865  \\
    \end{tabular}
  \end{center}
\end{table}
}
